---
title: "SIT741 Week 4 Practical"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```


## Learning Goal
1. You can use plots to select a candidate distribution family.
2. You can fit the maximum likelihood estimation for a univariate distribution.
3. You can use plots to check the fit of a distribution.


## Exploratory data analysis

We will use the *groundbeef* dataset from the *fitdistrplus* package. 

```{r}
library(tidyverse)
library(fitdistrplus)
data("groundbeef")
groundbeef
```

> Read the documentation for the dataset to gain an initial understanding of the data.

```{r}
groundbeef %>% 
  ggplot(aes(x = serving)) +
  geom_histogram()
hist(groundbeef[,1])
```

We can see some clumps of values. 
We can see it better with a QQ-plot, where the clumps are shown as flat line segments.
```{r}
groundbeef %>% 
  ggplot(aes(sample = serving)) +
  geom_qq()  #compare quantiles of your data with normal distribution
             ##longer horiontal values represents the pik values 
```
Or from the empirical cdf, where we see vertical jumps.
```{r}
groundbeef %>% 
  ggplot(aes(x = serving)) +
  stat_ecdf() ## the vertical long lines represent max probability , it also shows quantiles and the respectivw probabilitites 
```



The clumps can also be confirmed by counting observations by value.

```{r}
groundbeef %>% 
  count(serving) %>% 
  mutate(percent = n * 100 / sum(n)) %>% 
  arrange(desc(n)) ### 21 percent of data is 80.0 and it appears 54 times 
```

> Load the *toxocara* data from the *fitdistrplus* package. Explore the data using QQ-plot and the ECDF plot.

```{r}
data("toxocara")
  toxocara
```

## Choosing a candidate distribution family

To choose a candidate distribution family, 
we will use the Cullen and Frey graph, which is based on the idea of matching skewness and kurtosis.
(See Week 3 lecture for if you are unfamiliar with these concepts.)
The Cullen and Frey graph is implemented in the *descdist* function in *fitdistrplus*.

```{r}
library(magrittr)
groundbeef %$% 
  descdist(serving, boot = 100) ##take 100 samples 
```

We can see the distribution skew to the right and the kurtosis is greater than 3.

From the plot, we can exclude the uniform, the logistic, and the exponential distribution.

We will try the lognormal distribution. 

If we simulate a lognormal sample that resembles the original data, we can see how the Cullen and Frey graph differs.
```{r}
set.seed(1000)

simulated_data <- groundbeef %>% 
  mutate(logserv = log(serving)) %>% 
  mutate(simulated_serving = rlnorm(n = n(),
                                    meanlog = mean(logserv),
                                    sdlog = sd(logserv))) %>% 
  dplyr::select(-logserv)


simulated_data %>% 
  ggplot(aes(x = simulated_serving)) +
  geom_histogram()

simulated_data %>% 
  ggplot(aes(sample = simulated_serving)) +
  geom_qq()

simulated_data %$%  
  descdist(simulated_serving, boot = 100)
```

> Generate the Cullen and Frey graph for the *toxocara* data. Which distributions are not appropriate?

## Fit distribution via the maximum likelihood principle

The *fitdist* function in *fitdistrplus* implemented the maximum likelihood estimation.

```{r}
estimate <- groundbeef %$%
  fitdist(data = serving, 
          distr = "lnorm") ## fit to log normal distribution
```
From the plots below, we can see that the fit is poor near the right tail and also at the centre, corresponding to the clumps in the histogram.
```{r}
estimate %>% 
  plot
```

```{r}
estimate %>% summary
```

> Produce simulated data using the estimated parameter values. Then plot the QQ-plot and PP-plot. Do you see a lack of fit for the simulated data?


## Assess goodness of fit

We can use plots to assess the goodness of fit. 
The plots are similar to what we have seen before, the only difference being that we can choose to display the ones we believe relevant.

```{r}
estimate %>% 
  cdfcomp
```
```{r}
estimate %>% 
  denscomp()
```
```{r}
estimate %>% 
  qqcomp
```

```{r}
estimate %>% 
  ppcomp
```

Clearly the fit is rather poor. 


## Summary


The steps for fitting a distribution: 
 
* Explore data using histogram and QQ-plot
* Plot kurtosis against skewness to help choose a candidate distribution family
* Using maximum likelihood or other point estimation method to fit the distribution
* Assess the goodness of fit.


## Your turn

1. Fit a Gamma distribution on the beef serving data.
2. Fit a distribution on the *toxocara* data included in the *fitdistrplus* package.



