---
title: "SIT741 Week 7 Practical" 
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE)
```


## Learning Goal
1. You can use the *broom* R package to clean up model outputs.
2. You can perform model checking for linear regression through residuals.

## Credit rating vs income

We will use the Credit rating data from the *ISLR* package. 
Let's look at the data quickly using the *skim* function from the *skimr* package.
```{r}
library(tidyverse)
library(ISLR)
library(skimr)

skim(Credit)
```

Let's fit a linear model for the response variable *Rating*.
```{r}
Credit %>% 
  ggplot(aes(x = Income, 
             y = Rating)) +
  geom_point() +
  geom_smooth(method = 'lm')
```


```{r}
lmfit <- lm(Rating ~ Income, 
            data = Credit)

lmfit %>% 
  summary
```

## Clean up model outputs using *broom*

The *broom* package provides useful functions to clean up the outputs.

The *tidy* function outputs a data frame summarising the coefficients.
```{r}
library(broom)
lmfit %>% 
  tidy
```

The *augment* function outputs fitted values and residuals.
```{r}
(
results <- lmfit %>% 
  augment
)
```

The *glance* function outputs overall summaries of the fitted model.
```{r}
lmfit %>% 
  glance
```


## Checking distributional assumptions

We can use QQ-plot to check whether the residuals follow the theoretical distribution.

```{r}
results %>% 
  ggplot(aes(sample = .std.resid)) +
  geom_qq() + 
  geom_qq_line(col = 'steelblue')
```
The QQ-plot shows that the residuals have a finite range. This may indicate that a linear model is not  appropriate here. 


Then we should check that residuals are independent and having constant variance. 
This often involves plotting residuals against the fitted values and predictor values.

```{r}
results %>% 
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() + 
  geom_quantile() +
  geom_smooth(colour = 'firebrick') 
```

As the model has only one predictor, the following plot does not provide new information. 
```{r}
results %>% 
  ggplot(aes(x = Income, y = .resid)) +
  geom_point() + 
  geom_quantile() +
  geom_smooth(colour = 'firebrick') 
```

### Checking outliers and high leverage points

An observation has a high leverage if it has an unusual predictor value. 
This means potential risk of overfitting. We can check residuals against the leverage.

```{r}
results %>% 
  ggplot(aes(x = .hat, y = .resid)) +
  geom_point() + 
  geom_hline(yintercept = 0,
             linetype = 'dotted')
```
There are a couple of high leverage observations with large residuals.
Cook's distance measures the influence of each observation on the coefficients.
Plotting the leverage against the Cook's distance seems to suggest that these observations (most likely for high Income individuals) may need to be removed from the model.

```{r}
results %>% 
  mutate(big_hat = .hat > 0.03) %>% 
  ggplot(aes(x = .hat, y = .cooksd)) +
  geom_point(aes(color = big_hat)) + 
  geom_hline(yintercept = 0,
             linetype = 'dotted')
```

## Checking goodness of fit
The plots above can also show unexplained effects, indicating a lack of fit of the model.

A lack of fit can also be identified from summary statistics.
```{r}
lmfit %>% 
  glance
```

Here the model has a relatively low $R^2$. 

## Summary


1. Use *broom* package to process model outputs.
2. Check residuals for violation of modeling assumptions, and potential outliers and high leverage points causing overfitting risk. 
 


## Your turn

1. Try adding additional predictors in the model above and then analyse the results.



