---
title: "R Notebook"
output: html_notebook
---
################################################################################
################################################################################
Question 1: Import weather data 
################################################################################
################################################################################
1.1: No coding 
1.2 Import Data
```{r}
df_weather_mse <- read_csv("/Users/radhikaorari/Documents/UNI/23Semester2/SDA/Assignment2/weather_bom.csv")
```
```{r}
head(df_weather_mse)
```

Import all required libraries 
```{r}
library(tidyverse)
library(lubridate)
library(zoo)
```

Create Date column combining Day,Month,Year 
```{r}
df_weather_mse <- df_weather_mse %>%
  mutate(Date =  make_date(Year, Month, Day))
df_weather_mse
```

```{r}
# Extract the minimum and maximum dates from  data
min_date <- min(df_weather_mse$Date)
max_date <- max(df_weather_mse$Date)

# Convert minimum and maximum dates to human-readable format
min_date_readable <- format(min_date, "%B %d, %Y")
max_date_readable <- format(max_date, "%B %d, %Y")

# Print the human-readable dates
cat("Minimum Date :", min_date_readable, "\n")
cat("Maximum Date :", max_date_readable, "\n")

```

Check for missing values 
```{r}
missing_counts <- df_weather_mse %>%
  summarize_all(~sum(is.na(.)))
missing_counts
```
Function to replace missing values with a value before or after it.
```{r}
replace_na_with_avg <- function(x) {
  if (all(is.na(x))) {
    return(x)
  } else {
    return(zoo::na.approx(x))
  }
}
```

Replace the missing values 
```{r}
# List of columns to replace NAs
selected_columns <- c("TMAX", "TMIN", "Rainfall", "SunExposure")

# Replace NAs for the above columns using the across function
df_weather_mse <- df_weather_mse %>%
  mutate(across(all_of(selected_columns), replace_na_with_avg))

```

Recheck to ensure all missing values are treated 
```{r}
missing_counts <- df_weather_mse  %>%
  summarize_all(~sum(is.na(.)))
missing_counts
```
```{r}
df_weather_mse
```


```{r}
df_weather_mse %>%
  ggplot(aes(x = Date, y = TMAX)) + 
  geom_line()
```

```{r}
df_weather_mse %>%
  ggplot(aes(x = Date, y = TMIN)) + 
  geom_line()
```



```{r}
number_of_rows <- nrow(df_weather_mse)
print(number_of_rows)
head(df_weather_mse)

```
Question 2: Modeling:No code 

################################################################################
################################################################################
Question 3: Car accident data analysis for south eastern metropolitan Region
################################################################################
################################################################################
Q3.1 Which region do you pick?
```{r}
print(selected_region_data)
```
```{r}
selected_region_data$DATE_numeric <- as.numeric(as.Date(selected_region_data$DATE))
```

```{r}

selected_region_data %>% 
  ggplot(aes(x = DATE_numeric, 
             y = TOTAL_ACCIDENTS)) +
  geom_point() +
  geom_smooth(method = 'lm')
```
#####################################################################
3.2 Linear Model
#####################################################################
3.2.1 Fit the model  
```{r}
lmfit_car <- lm(TOTAL_ACCIDENTS ~ DATE_numeric, 
            data = selected_region_data)

lmfit_car %>% 
  summary
```
```{r}
library(broom)
lmfit_car %>% 
  tidy
```
```{r}
(
results_car <- lmfit_car %>% 
  augment
)
results_car
```
```{r}
lmfit_car %>% 
  glance
```

# Plot QQ plot 
```{r}
results_car %>% 
  ggplot(aes(sample = .std.resid)) +
  geom_qq() + 
  geom_qq_line(col = 'steelblue')
```


Plot Residuals vs fitted value:
```{r}
results_car %>% 
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() + 
  geom_quantile() +
  geom_smooth(colour = 'firebrick') 
```

Plot residual vs hat to analyse outliers and high leverage points
```{r}

results_car %>% 
  ggplot(aes(x = .hat, y = .resid)) +
  geom_point() + 
  geom_hline(yintercept = 0,
             colour = "red",
             linetype = 'dotted')
```

```{r}
results_car %>% 
  mutate(big_hat = .hat > 0.03) %>% 
  ggplot(aes(x = .hat, y = .cooksd)) +
  geom_point(aes(color = big_hat)) + 
  geom_hline(yintercept = 0,
             linetype = 'dotted')
```

################################################################################
3.3: Fitting GAM
################################################################################

Visualisation

```{r}
library(tidyverse)
library(ggplot2)
library(mgcv)

selected_region_data %>%
  ggplot(aes(x = DATE_numeric, y =TOTAL_ACCIDENTS)) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x), method.args = list(family =nb()))
```

Fit the GAM model
```{r}
library(mgcv)
gamfit1_car_nb <- gam(TOTAL_ACCIDENTS ~ s(DATE_numeric), 
                    data = selected_region_data, 
                    family = nb()
                    )
gamfit1_car_nb %>% 
  summary
```
```{r}
gam.check(gamfit1_car_nb)
```

```{r}
results_car_gam <- data.frame(
  .fitted = fitted(gamfit1_car_nb),
  .resid = resid(gamfit1_car_nb)
)

# Now, create the plot
library(ggplot2)
library(quantreg) # for geom_quantile

results_car_gam %>% 
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +  # Added alpha for better visualization if points overlap
  geom_quantile() +           # This will plot quantile regression lines
  geom_smooth(colour = 'firebrick', method = "loess") +  # Using loess for general trend
  theme_minimal() +  # Optional: Cleaner theme
  labs(title = "Fitted vs. Residuals",
       x = "Fitted Values",
       y = "Residuals")
```


```{r}
library(mgcv)
library(ggplot2)

# Extracting leverage (hat) values
hat_values <- influence(gamfit1_car_nb)

# Extracting residuals
residuals <- residuals(gamfit1_car_nb)

# Creating a data frame to hold these values
results_gam <- data.frame(hat = hat_values, resid = residuals)

# Creating the plot
results_gam %>%
  ggplot(aes(x = hat, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, colour = "red", linetype = 'dotted')

```
################################################################################
Q3.4: Augment
################################################################################
```{r}
selected_region_data$DayOfWeek <- factor(selected_region_data$DayOfWeek)
class(selected_region_data$DayOfWeek)
```

```{r}

gamfit2_car_nb_aug <- gam(TOTAL_ACCIDENTS ~ s(DATE_numeric) + s(DayOfWeek,k=7), 
             data = selected_region_data, 
             family = nb())
summary(gamfit2_car_nb_aug)
```


```{r}
gam.check(gamfit2_car_nb_aug)
```

```{r}
results_car_gam <- data.frame(
  .fitted = fitted(gamfit2_car_nb_aug),
  .resid = resid(gamfit2_car_nb_aug)
)

# Now, create the plot
library(ggplot2)
library(quantreg) # for geom_quantile

results_car_gam %>% 
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +  # Added alpha for better visualization if points overlap
  geom_quantile() +           # This will plot quantile regression lines
  geom_smooth(colour = 'firebrick', method = "loess") +  # Using loess for general trend
  theme_minimal() +  # Optional: Cleaner theme
  labs(title = "Fitted vs. Residuals",
       x = "Fitted Values",
       y = "Residuals")
```

################################################################################
Q3.5: AIC, coefficient estimates,graphs
################################################################################

```{r echo=FALSE}
AIC(lmfit_car,
    gamfit1_car_nb,
    gamfit2_car_nb_aug)
gam.check(gamfit2_car_nb_aug)
```
```{r}
summary(lmfit_car)
summary(gamfit1_car_nb)
summary(gamfit2_car_nb_aug)
```
```{r}
gam.check(gamfit2_car_nb_aug)
```

```{r}

gam.check(gamfit1_car_nb)
```


################################################################################
3.6:Residual analysis 
################################################################################
```{r}

# Obtain residuals for model 
lm_residuals <- residuals(lmfit_car)
gam_residuals1 <- residuals(gamfit1_car_nb)
gam_residuals2 <- residuals(gamfit2_car_nb_aug)


# Check for autocorrelation in residuals
acf(lm_residuals)
acf(gam_residuals1)
acf(gam_residuals2)



```
```{r}
head(selected_region_data)
```



################################################################################
################################################################################
Q4: Weather Data
################################################################################
################################################################################
Calculate temperature as mean of TMIN and TMAX
```{r}
# Create Temperature column as the average of TMIN and TMAX
df_weather_mse <- df_weather_mse %>%
  mutate(Temperature = (TMIN + TMAX) / 2)
```


Find T95 
```{r}
# Compute T95 as the 95th percentile of temperature
T95 <- quantile(df_weather_mse$Temperature, 0.95)
T95
```

Extract temperature for next 2 days 
```{r}
# Add lead (similar to shift in Python) of Temperature for the next days
df1_weather_mse <- df_weather_mse %>%
  mutate(Temperature_lead1 = lead(Temperature, 1),
         Temperature_lead2 = lead(Temperature, 2))
```

Calculate EHIsig
```{r}
# Calculate EHIsig
df1_weather_mse <- df1_weather_mse %>%
  mutate(EHIsig = (Temperature + Temperature_lead1 + Temperature_lead2) / 3 - T95)
```

Calculate EHIaccl
```{r}
# Calculate EHIaccl
df1_weather_mse <- df1_weather_mse %>%
  mutate(avg_prev_30 = rollapply(Temperature, width = 30, FUN = mean, fill = NA, align = "right")) %>%
  mutate(EHIaccl = (Temperature + Temperature_lead1 + Temperature_lead2) / 3 - avg_prev_30)
```

Calculate EHF
```{r}
# Calculate EHF
df1_weather_mse <- df1_weather_mse %>%
  mutate(EHF = pmax(0,EHIsig) * pmax(1, EHIaccl))
```

```{r}
print(df1_weather_mse )
```

Now we will align the dates as per car accident data.
```{r}
# Filter the data
df2_weather_mse <- df1_weather_mse %>% 
  filter(!(Date <= "2015-12-31" | Date >= "2020-07-01"))
```

Visualise headwave trend over the years
```{r}
df2_weather_mse %>% 
  ggplot() +
  geom_line(aes(Date, EHF))+
  labs(title = "EHF Over Time", x = "Years", y = "EHF")
```

```{r}
df2_weather_mse %>%
  mutate(MonthLabel = factor(Month, labels = month.abb)) %>% 
  ggplot() +
  geom_line(aes(x = MonthLabel, y = EHF)) +
  labs(title = "EHF Over The months", x = "Month", y = "EHF") 
```


Merge data
```{r}
df_merged_data <- merge(selected_region_data, df2_weather_mse[, c("DATE_numeric", "EHF")], by = "DATE_numeric", all.x = TRUE)
```

Fit new model with EHF
```{r}
gamfit3_CWehf <- gam(TOTAL_ACCIDENTS ~ s(DATE_numeric) + s(DayOfWeek, k=7) + s(EHF), 
                          data = df_merged_data, 
                          family = nb())
summary(gamfit3_CWehf)
```
4.3.2:Incorporate Sun Exposure

```{r}
df_merged_data1 <- merge(df_merged_data, df2_weather_mse[, c("DATE_numeric", "SunExposure")], by = "DATE_numeric", all.x = TRUE)
```

```{r}
gamfit4_CWsun <- gam(TOTAL_ACCIDENTS ~ s(DATE_numeric) + s(DayOfWeek, k=7) + s(SunExposure), 
                          data = df_merged_data1, 
                          family = nb())
summary(gamfit4_CWsun)
```
 
```{r echo=FALSE}
AIC(gamfit1_car_nb,
    gamfit2_car_nb_aug,
    gamfit3_CWehf,
    gamfit4_CWsun)
gam.check(gamfit4_CWsun)
```



